---
title: "Customer Segmentation for Tiered Pricing"
author: "Ife Abe"
date: "2024-10-29"
output: html_document
---

```{r setup, include=FALSE}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(cluster)    # For clustering
library(factoextra) # For visualization of clusters
set.seed(42) # Set seed for reproducibility
```

## 1. Generating Synthetic Customer Data

```{r generate-data}
# Number of customers
n <- 500 

# Generate synthetic data
customer_data <- data.frame(
  customer_id = 1:n,
  spending = rnorm(n, mean = 500, sd = 150),       # Monthly spending in dollars
  business_size = sample(1:3, n, replace = TRUE),   # 1: Small, 2: Medium, 3: Large
  usage_frequency = rpois(n, lambda = 10),          # Number of interactions per month
  satisfaction_score = runif(n, min = 1, max = 5)   # Customer satisfaction score (1-5)
)

# Ensure no negative spending values
customer_data <- customer_data %>%
  mutate(spending = ifelse(spending < 0, 0, spending))

# Display the first few rows of data
head(customer_data)
```

## 2. Preprocessing and Feature Scaling

```{r scaling-data}
# Scale the numerical features for clustering
customer_data_scaled <- customer_data %>%
  select(spending, business_size, usage_frequency, satisfaction_score) %>%
  scale()
```

## 3. Finding the Optimal Number of Clusters

```{r elbow-method}
# Function to calculate total within-cluster sum of square
wss <- function(k) {
  kmeans(customer_data_scaled, k, nstart = 10)$tot.withinss
}

# Compute and plot WSS for k = 1 to k = 10
k_values <- 1:10
wss_values <- sapply(k_values, wss)

plot(k_values, wss_values, type = "b", pch = 19, frame = FALSE,
     xlab = "Number of Clusters K",
     ylab = "Total Within-Clusters Sum of Squares",
     main = "Elbow Method for Optimal K")
```

## 4. Applying K-Means Clustering

Based on the elbow plot, we’ll use `k = 3` clusters.

```{r clustering}
# Apply k-means clustering with chosen number of clusters (e.g., k = 3)
k <- 3
kmeans_result <- kmeans(customer_data_scaled, centers = k, nstart = 10)
customer_data$cluster <- as.factor(kmeans_result$cluster)
```

## 5. Visualizing Clusters

```{r cluster-visualization}
# Plot clusters using factoextra package
fviz_cluster(kmeans_result, data = customer_data_scaled,
             geom = "point", ellipse.type = "convex", 
             ggtheme = theme_minimal(),
             main = "K-means Clustering of Customer Data")
```

### Spending vs Usage Frequency with Clusters

```{r spending-usage-plot}
# Visualize clusters based on spending and usage frequency
ggplot(customer_data, aes(x = spending, y = usage_frequency, color = cluster)) +
  geom_point(alpha = 0.7) +
  labs(title = "Customer Segmentation by Spending and Usage Frequency",
       x = "Monthly Spending (USD)",
       y = "Usage Frequency") +
  theme_minimal()
```

## 6. Analyzing Cluster Profiles

```{r cluster-profiles}
# Summary of each cluster
cluster_summary <- customer_data %>%
  group_by(cluster) %>%
  summarize(
    avg_spending = mean(spending),
    avg_business_size = mean(business_size),
    avg_usage = mean(usage_frequency),
    avg_satisfaction = mean(satisfaction_score)
  )
cluster_summary
```

## 7. Pricing Tier Recommendations

Using the clusters, we can recommend pricing tiers based on the segment’s average characteristics.

```{r pricing-recommendations}
# Example pricing recommendations for each cluster
tier_recommendations <- data.frame(
  cluster = 1:k,
  description = c("Small Business Tier", "Medium Business Tier", "Enterprise Tier"),
  suggested_price = c(300, 500, 700) # Hypothetical prices for each segment
)
tier_recommendations
```

## Conclusion

This analysis provides a basic framework for customer segmentation using k-means clustering. Based on the cluster characteristics, we can tailor pricing tiers that align with each segment's willingness to pay and product usage, enhancing revenue potential and customer satisfaction.
