---
title: "Customer Segmentation for Tiered Pricing"
author: "Ife Abe"
date: "2024-10-29"
output: html_document
---

```{r setup, include=FALSE}

# Load necessary libraries

library(dplyr)
library(ggplot2)
library(cluster)    
library(factoextra) 

set.seed(42) 

```

## 1. Generating Synthetic Customer Data

```{r generate-data}

# Number of customers

n <- 500 

# Generate synthetic data

customer_data <- data.frame(
  customer_id = 1:n,
  spending = rnorm(n, mean = 500, sd = 150),       
  business_size = sample(1:3, n, replace = TRUE),   
  usage_frequency = rpois(n, lambda = 10),         
  satisfaction_score = runif(n, min = 1, max = 5)   
)

# Ensure no negative spending values

customer_data <- customer_data %>%
  mutate(spending = ifelse(spending < 0, 0, spending))

# Display the first few rows of data

head(customer_data)

```

## 2. Preprocessing and Feature Scaling

```{r scaling-data}

# Scale the numerical features for clustering

customer_data_scaled <- customer_data %>%
  select(spending, business_size, usage_frequency, satisfaction_score) %>%
  scale()

```

## 3. Finding the Optimal Number of Clusters

```{r elbow-method}

# Function to calculate total within-cluster sum of square

wss <- function(k) {
  kmeans(customer_data_scaled, k, nstart = 10)$tot.withinss
}

# Compute and plot WSS for k = 1 to k = 10

k_values <- 1:10
wss_values <- sapply(k_values, wss)

plot(k_values, wss_values, type = "b", pch = 19, frame = FALSE,
     xlab = "Number of Clusters K",
     ylab = "Total Within-Clusters Sum of Squares",
     main = "Elbow Method for Optimal K")

```

## 4. Applying K-Means Clustering

Based on the elbow plot, we’ll use `k = 3` clusters.

```{r clustering}

# Apply k-means clustering with chosen number of clusters 

k <- 3
kmeans_result <- kmeans(customer_data_scaled, centers = k, nstart = 10)
customer_data$cluster <- as.factor(kmeans_result$cluster)

```

## 5. Visualizing Clusters

```{r cluster-visualization}

# Plot clusters using factoextra package

fviz_cluster(kmeans_result, data = customer_data_scaled,
             geom = "point", ellipse.type = "convex", 
             ggtheme = theme_minimal(),
             main = "K-means Clustering of Customer Data")

```

### Spending vs Usage Frequency with Clusters

```{r spending-usage-plot}

# Visualize clusters based on spending and usage frequency

ggplot(customer_data, aes(x = spending, y = usage_frequency, color = cluster)) +
  geom_point(alpha = 0.7) +
  labs(title = "Customer Segmentation by Spending and Usage Frequency",
       x = "Monthly Spending (USD)",
       y = "Usage Frequency") +
  theme_minimal()

```

## 6. Analyzing Cluster Profiles

```{r cluster-profiles}

# Summary of each 

cluster_summary <- customer_data %>%
  group_by(cluster) %>%
  summarize(
    avg_spending = mean(spending),
    avg_business_size = mean(business_size),
    avg_usage = mean(usage_frequency),
    avg_satisfaction = mean(satisfaction_score)
  )
cluster_summary

```

## Interpretation

Cluster 1 includes small and medium businesses that spend the most but are least satisfied, meaning they are high-value but at risk.
Cluster 2 includes small businesses with the highest usage and satisfaction, representing loyal and engaged users.
Cluster 3 includes the smallest businesses with moderate spending and satisfaction, making them stable but lower-growth users.

Overall, retain Cluster 1, reward Cluster 2, and develop Cluster 3.

## 7. Pricing Tier Recommendations

Using the clusters, we can recommend pricing tiers based on the segment’s average characteristics.

```{r pricing-recommendations}

# Example pricing recommendations for each cluster

tier_recommendations <- data.frame(
  cluster = 1:k,
  description = c("Small Business Tier", "Medium Business Tier", "Enterprise Tier"),
  suggested_price = c(300, 500, 700) # Hypothetical prices for each segment
)
tier_recommendations

```

## Conclusion

This analysis demonstrates how k-means clustering can segment customers into meaningful groups based on spending, business size, usage, and satisfaction. The resulting segments naturally align with three pricing tiers, small business ($300), medium business ($500), and enterprise ($700). By matching pricing and engagement strategies to each group’s characteristics, the business can improve revenue efficiency, retention, and overall customer satisfaction.
